D.I.A.G.R.A.M.: Development of Image Analysis for Graph Recognition And Modeling

### 2025-06-29 doc/article/article.tex

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{cite}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\subsection{Dataset Utilization Strategy}

We implemented a targeted approach for dataset utilization based on the specific requirements of our two-stage architecture:

@@

\textbf{Extractor Training:} For the specialized extraction tasks, we focused exclusively on flowchart and graph diagrams from the FCB, OHFCD\_1, and Finite Automata datasets. This targeted approach allows the extraction models to achieve high precision on the specific structural elements relevant to these diagram types.

@@

\subsection{Object Detection}
\label{sec:obj_det}
TODO: Talk about the net.

@@

\section{Discussion}
TODO

>>>>

\usepackage{graphicx}
\usepackage{multirow}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{cite}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

Many datasets don't include the task at hand - which is focused on recognizing flowchart and graph diagrams - but could be in future works. Moreover, many datasets not apt to our task were used to further enhance the capability of the classifier - letting it know if an uploaded image couldn't be further processed.

\subsection{Dataset Usage Strategy}

We implemented a targeted approach for dataset utilization based on the specific requirements of our architecture:

@@

\textbf{Extractor Training:} For the specialized extraction tasks, we focused exclusively on the first three datasets: flowchart and graph diagrams from the FCB, OHFCD\_1, and Finite Automata datasets. This targeted approach allows the extraction models to achieve high precision on the specific structural elements relevant to these diagram types, which were the core of our task.

@@

\subsection{Object Detection}
\label{sec:obj_det}

For bounding box detection of diagram elements, we employed a Faster R-CNN architecture \cite{rcnn} with a ResNet-50 Feature Pyramid Network (FPN) backbone \cite{lin2017feature}. This two-stage object detection framework provides robust performance for multi-class detection tasks while maintaining computational efficiency suitable for diagram analysis applications.
Its main components are:

\textbf{Backbone Network:} The ResNet-50 backbone serves as the feature extraction component, providing representations through its four residual blocks.

\textbf{Feature Pyramid Network:} The FPN enhancement augments the ResNet-50 backbone features by creating a top-down pathway with connections.

\textbf{Region Proposal Network (RPN):} The RPN generates object proposals by sliding a small network over the feature maps from the FPN. At each spatial location, the RPN predicts object presence probability and bounding box refinements for different scales and aspect ratios.

\textbf{ROI Head:} The second stage performs precise classification and bounding box regression on the proposals generated by the RPN. Feature vectors are extracted using ROI Align operations. \\

The original model was trained over the COCO dataset \cite{coco}, which comprised 91 classes. It should be highlighted that the original training dataset was not in any way related to diagrams: with an efficient finetune, the network was able to swiftly change to our task. \\

@@

\section{Discussion}

The development of D.I.A.G.R.A.M. offers an insightful perspective on the integration of deterministic logic and deep learning for the interpretation of handwritten diagrams. One of the most valuable outcomes of this project is the successful design of a complete and functional pipeline (see Figure~\ref{fig:results}), which transforms raw visual input into a structured, editable format. This end-to-end workflow—from classification to rendering—highlights the feasibility of modular architectures for diagram understanding.

\begin{figure}[htbp]
	\centering

	% Row 1
	\begin{subfigure}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{ex1.png}
		\caption{An easy flowchart diagram.}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.4\linewidth}
		\includegraphics[width=40pt, height=150pt]{ex2.png}
		\caption{The network's output.}
	\end{subfigure}

	\vspace{1em} % spacing between rows

	% Row 2
	\begin{subfigure}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{ex3.png}
		\caption{A complex graph diagram.}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\linewidth}
		\includegraphics[width=80pt, height=150pt]{ex4.png}
		\caption{The network's output.}
	\end{subfigure}

	\vspace{1em}

	% Row 3
	\begin{subfigure}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{ex5.png}
		\caption{A complex flowchart diagram.}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\linewidth}
		\includegraphics[width=\linewidth]{ex6.png}
		\caption{The network's output.}
	\end{subfigure}

	\caption{A few results of our network.}
	\label{fig:results}
\end{figure}

Among the modules, the Extractor proved particularly effective. Leveraging deep neural networks, it reliably identified diagram elements across diverse samples, validating the suitability of data-driven approaches for this subtask. Conversely, the Classifier, while operational, showed room for improvement, particularly in edge cases or less represented diagram categories. This suggests that the current training set may lack the diversity required for robust generalization, pointing to data augmentation and dataset expansion as immediate avenues for refinement.

The system as a whole adheres to principles of scalability and generality. By decoupling the diagram type detection from the extraction and compilation logic, D.I.A.G.R.A.M. maintains flexibility in handling new diagram formats with minimal architectural changes. However, its performance remains sensitive to input quality and assumes a level of neatness in the handwritten diagrams that may not always be guaranteed in real-world scenarios.

In comparison to related work, D.I.A.G.R.A.M. distinguishes itself by targeting offline, image-based recognition (as opposed to stroke-based online methods) while retaining support for structural output generation. This positions it as a hybrid between traditional rule-based recognizers and modern neural approaches.
